### 2012-12-14 doc/description.tex
formating improvement, mappings improvement

$$
Application = (Class*)
$$
$classes : Application \rightarrow Class*$ \\
$classes((classes)) = classes$

@@

$$Class = (label, Property*, Association*)$$
$name : Class \rightarrow Label$ \\
$name((lab, props, assocs)) = lab$ \\ \\
$properties : Class \rightarrow Property*$ \\
$properties((lab, props, assocs)) = props $\\ \\
$primitiveProperties : Class \rightarrow Property*$ \\
$primitiveProperties((lab, props, assocs)) = props', \forall p \in props' : cardinality(p) \leq 1 $\\ \\
$collectionProperties : Class \rightarrow Property*$ \\
$collectionProperties((lab, props, assocs)) = props', \forall p \in props' : cardinality(p) > 1 $\\ \\
$associations : Class \rightarrow Association*$ \\
$associations((lab, props, assocs)) = assocs $

\paragraph{Property} represents a feature of  a class which is can be represented as a primitive type. The property can be mandatory, can have a default value.
$$
Property = (label, AppType, DefaultValue, Cardinality)
$$
$name : Property \rightarrow Label$ \\
$name((lab, t, val, n)) = lab$ \\ \\
$type : Property \times AppType$ \\
$type((lab, t, val, n)) = t$ \\ \\
$cardinality : Property \rightarrow NzNat$ \\
$cardinality((lab, t, val, n)) = n$ \\ \\
$owningClass : Property \times Application \rightarrow Class $ \\
$owningClass(p, a) = c, c \in classes(a) \wedge p \in properties(c) $

\paragraph {Association} represents a connection between two classes. It has a unique name and it contains a name of the class which is referenced by the association. The class which owns the association is consider to be a starting class of an association, referenced class is consider to be an ending class of an association. The cardinalities defines the multiplicity of the association.
$$
Association = (label, classRef, startCardinality, endCardinality)
$$
$name : Association \rightarrow Label$ \\
$name((lab, ref, n_1, n_2)) = lab$\\ \\
$reference : Association \rightarrow Label$ \\
$reference((lab, ref, n_1, n_2)) = ref$\\ \\
$startCardinality : Association \rightarrow NzNat$ \\
$startCardinality((lab, ref, n_1, n_2)) = n_1$\\ \\
$endCardinality : Association \rightarrow NzNat$ \\
$endCardinality((lab, ref, n_1, n_2)) = n_2$ \\ \\
$owningClass : Association \times Application \rightarrow Class $ \\
$owningClass(as, a) = c, c \in classes(a) \wedge as \in associations(c) $

@@

$
AddProperty(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \not \exist q \in properties(p) : name(q) \neq name(p)}{
 properties(c) \cup p }
$
$$AddAssociation : Label \times Association \times Application \rightarrow Application $$
$
AddAssociation(l, s, a) = \inference{\not \exists f \in associations(a) : name(f) = name(s) \\ \wedge \exists c in classes(a) : name(c) = l \\ \wedge \exists k \in classes(a) : name(k) = reference(s) }{associations(a) \cup s}
$

\subsection{Application Deconstruction}
$$RemoveClass: Label \times Application \rightarrow Application $$
$
RemoveClass(q, a) = \inference{\exists c \in classes(a) : name(c) = q \\
\wedge not$ $instantiated(c) = \{ \} \wedge not$ $associated(c)}{classes(a) \setminus c }
$

@@

\section{Database model}
The relational database consists of database schema which defines the structure of the database and data. The database is a pair od sets:
$$
Database = (table*, data*)
$$

\subsection{Database Schema Model}
The database schema model is defined by following concepts:
\begin{itemize}
	\item Table: table represents a basic concept of database schema. It has a unique name, one or more column and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
$$
Table = (label, primaryKey, Column*, ForeignKey*)
$$
	\item Column : columns define possible data values and types which can be part of a table record.
$$
Column = (label, Db-Type, Constraint*)
$$

	\item Foreign key:
$$
ForeignKey = (label, tableRef, Constraint*)
$$

	\item Primary key:
$$
PrimaryKey =  (label) 	
$$

	\item Db-Type: represents primitive types in the database. There are usually defined types such as Char, Integer, Boolean etc. There is only one type defined in the model which is called DB-STRING.
$$
Db-Type = DBSTRING
$$

	\item Constraint: there are two types of constraints defined in the model. Both constraints are column constraints - first constraint defines unique records in a column, second constraint defines non-empty columns.
$$
Constraint = NOTNULL | UNIQUE 
$$
\end{itemize}

@@

\subsection{Mapping of Types}
The mapping of types assumes a bijection between application types and database types, otherwise these has to be additional information for type mapping. We focus on structural and data change therefore we simplify types and its mapping.
$$
\rho : AppType \rightarrow DbType
$$

$$
\rho(APPSTRING) = DBSTRING
$$

@@

$$
\rho : Class \rightarrow Table
$$

$$
\rho(c) = table(name(c), primaryKey("Id"), \rho(properties(c), \{ \})
$$

@@

$\rho(p, d) = \begin{cases}
 \inference{cardinality(p) = 0 \\ \wedge \rho(owningClass(p) \in classes(d)}{columns(owninngClass(p)) \cup column(name(p), \rho(type(p)), \{ \})} \\ \\

 \inference{cardinality(p) = 1 \\ \wedge \rho(owningClass(p) \in classes(d)}{columns(owninngClass(p)) \cup col(name(p), \rho(type(p)), NOTNULL)} \\ \\

 \inference{cardinality(p) > 2 \\ \wedge \rho(owningClass(p) \in classes(d)}{\begin{gathered} tables(d) \cup table(name(p), pk("Id"), col("value", DBSTRING), \\ fk(name(p), \rho(owningClass(p))) \end{gathered}}
 \end{cases}$

@@

$
\rho(a,d) = \begin{cases}
 \inference{startCardinality(a) = 0 \\ \wedge \rho(owningClass(p) \in classes(d)  \wedge \rho(reference(a)) \in tables(d)}{foreignKeys(\rho(assoc(a)) \cup foreignKey(name(a), owningClass(a),  \{ \}) } \\ \\

 \inference{startCardinality(a) = 1 \\ \wedge \rho(owningClass(p) \in tables(d)  \wedge \rho(reference(a)) \in tables(d)}{\begin{gathered}
 foreignKeys(\rho(assoc(a)) \cup foreignKey(name(a), \\ owningClass(a),  NOTNULL)
\end{gathered}
} \\ \\

  \inference{startCardinality > 1 \\ \wedge \rho(owningClass(p)) \in tables(d) \wedge \rho(reference(a)) \in tables(d)}{
  \begin{gathered}
  tables(d) \cup table(name(a), pk("Id"), \{ \}, \\ foreignKey(owningClass(a),owningClass(a), NOTNULL) \\ foreignkey(reference(a), reference(a), NOTNULL)) 
  \end{gathered}}

@@

\section{Data Evolution in ORM System}
The Object-Relational system consist of an application and a database. The system has to be in consistent state, therefore the evolution of application has to be propagated correctly into database. Each evolution operation consist of two parts - applying evolution operation on application and on database. In general an evolutionary operation $X : OrmSystem \rightarrow OrmSystem $ can be defined as:

$$
\inference[X(s) = ]{}{system(X(application(s)), \rho_{X}(database(s))) }
$$
where the $\rho_X$ is the mapping of the evolution from application to database. The mapping of the evolution change is called extended Object-Relational Mapping (eORM). The eORM mapping helps to assure consistency of the system by transforming application evolution into database schema and data. Each eORM function applied on an ORM system produces a new generation of the system. If the function cannot be applied on a system the system cannot be evolved by the function.

@@

$
\rho_{addClass}(c, d) = \inference{\forall t \in tables(d) : name(t) \neq name(\rho(c)) \\
	\forall a \in \rho(associations(c)) : reference(a) \in tables(d) 
}{ %TODO podmínky?
\begin{gathered}
tables(d) \cup \rho(c) \cup \rho(properties(c)) \cup \rho(associations(c))
\end{gathered}
}
$
\subsubsection{Add Property}
$
\rho_{addProperty}(l, p, d) = 
\inference{\exists t \in tables(d) : name(t) = l}{\rho(p,d)} 
$
\subsection{Add Association}
$
\rho_{AddAssociation}(l, a, d) = \inference{\exists t_1, t_2 \in tables(d) : name(t_1) = l \\ \wedge name(t_2) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_1) : name(f) = name(a) }{\rho(a,d)}
$

\subsection{Deconstruction of a System}
\subsubsection{Remove Class}
$
\rho_{removeClass}(c,d) = \inference{\rho(c) \in tables(d) \\ \wedge \not \exists r \in data(d) : owningTable(r) \neq \rho(c) \\ \not \exists j \in tables(d), f \in foreignKeys(j) : reference(f) = \rho(c)  }{tables(d) \setminus \rho(c)}
$
%TODO remove by label

\subsubsection{Remove Property}
$
\rho_{removeProperty}(l_c, l_p, d) = \inference{\exists t \in tables(d) : name(t) = l_c \\ \wedge \exists c \in columns(t) : name(c) = l_p  }{
columns(t) \setminus c
}
$
%TODO 1:N M:N

>>>>

$$
Application = (Class*, Instance*)
$$
$classes : Application \rightarrow Class*$ \\
$classes((classes)) = classes$ \\ \\
$instantiated : Class \times Application \rightarrow Boolean $
%TODO: variable types

@@

$$Class = (label, Property*, Association*)$$
$name : Class \rightarrow Label$ \\
$name((lab, props, assocs)) = lab$ \\ \\
$properties : Class \rightarrow Property*$ \\
$properties((lab, props, assocs)) = props $\\ \\
$primitiveProperties : Class \rightarrow Property*$ \\
$primitiveProperties((lab, props, assocs)) = props', \forall p \in props' : cardinality(p) \leq 1 $\\ \\
$collectionProperties : Class \rightarrow Property*$ \\
$collectionProperties((lab, props, assocs)) = props', \forall p \in props' : cardinality(p) > 1 $\\ \\
$associations : Class \rightarrow Association*$ \\
$associations((lab, props, assocs)) = assocs $ \\ \\
$associated :  Class \times Application \rightarrow Boolean $
%TODO: variable types

\paragraph{Property} represents a feature of  a class which is can be represented as a primitive type. The property can be mandatory, can have a default value.
$$
Property = (label, AppType, DefaultValue, Cardinality)
$$
$name : Property \rightarrow Label$ \\
$name((lab, t, val, n)) = lab$ \\ \\
$type : Property \times AppType$ \\
$type((lab, t, val, n)) = t$ \\ \\
$cardinality : Property \rightarrow NzNat$ \\
$cardinality((lab, t, val, n)) = n$ \\ \\
$owningClass : Property \times Application \rightarrow Class $ \\
$owningClass(p, a) = c, c \in classes(a) \wedge p \in properties(c) $
%TODO: variable types

\paragraph {Association} represents a connection between two classes. It has a unique name and it contains a name of the class which is referenced by the association. The class which owns the association is consider to be a starting class of an association, referenced class is consider to be an ending class of an association. The cardinalities defines the multiplicity of the association.
$$
Association = (label, classRef, startCardinality, endCardinality)
$$
$name : Association \rightarrow Label$ \\
$name((lab, ref, n_1, n_2)) = lab$\\ \\
$reference : Association \rightarrow Label$ \\
$reference((lab, ref, n_1, n_2)) = ref$\\ \\
$startCardinality : Association \rightarrow NzNat$ \\
$startCardinality((lab, ref, n_1, n_2)) = n_1$\\ \\
$endCardinality : Association \rightarrow NzNat$ \\
$endCardinality((lab, ref, n_1, n_2)) = n_2$ \\ \\
$owningClass : Association \times Application \rightarrow Class $ \\
$owningClass(as, a) = c, c \in classes(a) \wedge as \in associations(c) $
%TODO: variable types

@@

$
AddProperty(l, p, a) = \inference{\exists c \in classes(a) : name(c) = l \\ \wedge \forall q \in properties(p) : name(q) \neq name(p)}{
 properties(c) \cup p }
$
$$AddAssociation : Label \times Association \times Application \rightarrow Application $$
$
AddAssociation(l, s, a) = \inference{\forall f \in associations(a) : name(f) \neq name(s) \\ \wedge \exists c \in classes(a) : name(c) = l \\ \wedge \exists k \in classes(a) : name(k) = reference(s) }{associations(a) \cup s}
$

\subsection{Application Deconstruction}
$$RemoveClass: Label \times Application \rightarrow Application $$
$
RemoveClass(q, a) = \inference{\exists c \in classes(a) : name(c) = q \\
\wedge not$ $instantiated(c) = \varnothing \wedge not$ $associated(c)}{classes(a) \setminus c }
$

@@

\section{Database Component Model}
The relational database consists of database schema which defines the structure of the database and data. The database is a pair od sets:
$$
Database = (table*, data*)
$$

\subsection{Database Schema Model}
The database schema model is defined by following concepts:
\paragraph{Table} represents a basic concept of database schema. It has a unique name, one or more column and it can be related to other tables in the schema by foreign keys. Rows in the table represents stored data.
$$
Table = (label, primaryKey, Column*, ForeignKey*)
$$
$name : Table \rightarrow Label $ \\
$name((lab, pk, cols, fks)) = lab$ \\ \\
$primaryKey : Table \rightarrow PrimaryKey $ \\
$primaryKey((lab, pk, cols, fks)) = pk$ \\ \\
$columns : Table \rightarrow Column* $ \\
$columns((lab, pk, cols, fks)) = cols$ \\ \\
$foreignKeys : Table \rightarrow ForeignKey* $ \\
$foreignKeys((lab, pk, cols, fks)) = fks$

\paragraph{Column} columns define possible data values and types which can be part of a table record.
$$
Column = (label, DbType, DefaultValue, Constraint*)
$$
$name : Column \rightarrow Label $ \\
$name((lab, t, val, cons)) = lab $ \\ \\
$type : Column \rightarrow DbType $ \\
$type((lab, t, val, cons)) = t $ \\ \\
$constraints : Column \rightarrow Constraint* $ \\
$constraints((lab, t, val, cons)) = cons $ \\ \\
$owningTable : Column \times Database \rightarrow Table $ \\
$constraints(c, d) = tab, tab \in tables(d) \wedge c \in columns(d) $

\paragraph{Foreign key}
$$
ForeignKey = (label, tableRef, Constraint*)
$$
$name : ForeignKey \rightarrow Label $ \\
$name((lab, tRef, cons)) = lab $ \\ \\
$reference : ForeignKey \rightarrow Label $ \\
$reference((lab, tRef, cons)) = tRef $ \\ \\
$reference : ForeignKey \rightarrow Constraint* $ \\
$reference((lab, tRef, cons)) = cons $ \\ \\
$owningTable : ForeignKey \times Database \rightarrow Table $ \\
$owningTable(fk, d) = tab, tab \in tables(d) \wedge fk \in foreignKeys(d) $

\paragraph{Primary key}
$$
PrimaryKey =  (label) 	
$$

\paragraph{DbType} represents primitive types in the database. There are usually defined types such as Char, Integer, Boolean etc. There is only one type defined in the model which is called DBSTRING.
$$
DbType = DBSTRING
$$

\paragraph{Constraint} there are two types of constraints defined in the model. Both constraints are column constraints - first constraint defines unique records in a column, second constraint defines non-empty columns.

\begin{center}
$Constraint = NOTNULL$ $|$ $UNIQUE $
\end{center}

@@

\subsection{Mapping of Types}
The mapping of types assumes a bijection between application types and database types, otherwise these has to be additional information for type mapping. We focus on structural and data change therefore we simplify types and its mapping.

\begin{center}
$ \rho : AppType \rightarrow DbType $ \\
$ \rho(APPSTRING) = DBSTRING $
\end{center}

@@

\begin{center}
$\rho : Class \rightarrow Table $ \\ 
$ \rho(c) = table(name(c), primaryKey("Id"), \rho(primitiveProperties(c), \varnothing $\\$ \cup \rho(associations(c)) \cup \rho(collectionProperties)$
\end{center}

@@

$\rho(p, d) = \begin{cases}
 \inference{cardinality(p) = 0 \\ \wedge \rho(owningClass(p)) \in classes(d)}{columns(\rho(owningClass(p))) \cup column(name(p), \rho(type(p)), \varnothing)} \\ \\

 \inference{cardinality(p) = 1 \\ \wedge \rho(owningClass(p) \in classes(d)}{columns(\rho(owningClass(p))) \cup column(name(p), \rho(type(p)), NOTNULL)} \\ \\

 \inference{cardinality(p) > 2 \\ \wedge \rho(owningClass(p) \in classes(d)}{\begin{gathered} table(name(p), primaryKey("Id"), column("value", \rho(type(p))), \\ foreignKey(name(p), name(owningClass(p))) \end{gathered}}
 \end{cases}$

@@

$
\rho(a,d) = \begin{cases}
 \inference{startCardinality(a) = 0 \\ \wedge \rho(owningClass(p)) \in classes(d)  \wedge \rho(reference(a)) \in tables(d)}{
 \begin{gathered}
 foreignKeys(\rho(reference(a))) \\ \cup foreignKey(name(a), owningClass(a),  \varnothing) 
 \end{gathered}
 }

  \\ \\

 \inference{startCardinality(a) = 1 \\ \wedge \rho(owningClass(p) \in tables(d)  \wedge \rho(reference(a)) \in tables(d)}{\begin{gathered}
foreignKeys(\rho(reference(a))) \\ \cup  foreignKey(name(a), \\ owningClass(a),  NOTNULL)
\end{gathered}
} \\ \\

  \inference{startCardinality > 1 \\ \wedge \rho(owningClass(p)) \in tables(d) \wedge \rho(reference(a)) \in tables(d)}{
  \begin{gathered}
 table(name(a), pk("Id"), \varnothing, \\ foreignKey(owningClass(a),owningClass(a), NOTNULL) \\ foreignkey(reference(a), reference(a), NOTNULL)) 
  \end{gathered}}

@@

\section{Data Evolution in ORM System}
The Object-Relational system consist of an application and a database. The system has to be in consistent state, therefore the evolution of application has to be propagated correctly into database. Each evolution operation consist of two parts - applying evolution operation on application and on database. The mapping of the evolution change is called extended Object-Relational Mapping (eORM). The eORM mapping helps to assure consistency of the system by transforming application evolution into database schema and data. Each eORM function applied on an ORM system produces a new generation of the system. If the function cannot be applied on a system the system cannot be evolved by the function.

@@

$
\rho_{addClass}(c, d) = \inference{\forall t \in tables(d) : name(t) \neq name(\rho(c)) \\
	\forall a \in \rho(associations(c)) : reference(a) \in tables(d) 
}{
\begin{gathered}
tables(d) \cup \rho(c) \cup \rho(collectionProperties(c)) \\ \cup \rho(associations(c))
\end{gathered}
}
$
\subsubsection{Add Property}
$
\rho_{addProperty}(l, p, d) = 
\inference{\exists t \in tables(d) : name(t) = l}{\rho(p,d)} 
$
\subsection{Add Association}
$
\rho_{AddAssociation}(l, a, d) = \inference{\exists t_1, t_2 \in tables(d) : name(t_1) = l \\ \wedge name(t_2) = reference(a) \\ \wedge \not \exists f \in foreignKeys(t_1) : name(f) = name(a) }{\rho(a,d)}
$

\subsection{Deconstruction of a System}
\subsubsection{Remove Class}
$
\rho_{removeClass}(c,d) = \inference{\rho(c) \in tables(d) \\ \wedge \not \exists r \in data(d) : owningTable(r) = \rho(c) \\ \not \exists j \in tables(d), f \in foreignKeys(j) : reference(f) = \rho(c)  }{tables(d) \setminus \rho(c)}
$
%TODO remove by label

\subsubsection{Remove Property}
$
\rho_{removeProperty}(l_c, l_p, d) = \begin{cases}
 \inference{\exists t \in tables(d) : name(t) = l_c \\ \wedge \exists c \in columns(t) : name(c) = l_p  }{
columns(t) \setminus c
} \\ \\
  \inference{}{}
 \end{cases}

$
%TODO 1:N M:N
